Gap formation and stability in non-isothermal protoplanetary discs

### 2014-10-28 paper/results.tex
Updated Additional Analysis in results section

\section{Growth of non-axisymmetric modes without the influence of the
  planet}\label{linear1}
In this section, the planet is introduced at $t=20P_0$ and 
its potential switched on over $10P_0$. At $t=30P_0$ we switch off the
planet potential and azimuthally average the surface density, energy
and velocity fields. At this point the planet has carved a partial
gap. We then perturb the surface density for $r>r_p$ and continue to 
evolve the disc. We impose  sinusoidal perturbations with 
azimuthal wavenumbers $m\in[1,10]$. {\bf what are PERTMIN and PERTMAX values?}  
This procedure allows us to analyse the growth of 
non-axisymmetric modes associated with the gap, but without
complications from non-axisymmetry arising directly from disc-planet
interaction.

@@

Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
with $r_\mathrm{out}=25r_\mathrm{in}$, 
and we consider 
%for a time up to $200P_0$ allowing high resolution and detection of high
%$m$ structures.
cases of $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
and slowly cooled discs.  {\bf boundary condition for these sims?}

@@

@@

In all cases we observe growth of non-axisymmetric structures after the
planet potential is switched off and the system subject to random
perturbations. An example is shown in Fig. \ref{linearmodes} for 
$\tilde{\beta}=10$. We charcterize these
modes with an azimuthal wavenumber $m$ and growth rate $\gamma(m)$ as defined by
Eq.~\ref{fouriertransform}---\ref{growth}. Table \ref{modetable}
lists the growth rates for the three cooling
times, measured during linear growth.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
    0cm 0cm 0cm]{figures/linear_stability}
  \caption{Evolution of azimuthal Fourier modes of disc surface
    density, non-dimenionlized by the axisymmetric mode for the
    `planet-off' simulations with $\tilde{\beta}=10$. Colours correspond
    to different $m$ values. The $m=3$ component is the fastest growing
    mode during linear growth and has a corresponding
    $\gamma=0.017\Omega(r_p)$. {\bf is the $\Sigma_0$ at
      t=0?} \label{linearmodes}}  
\end{figure}

@@

\begin{figure*}
  \centering
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity_lowb}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity_medb}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity_highb}
  }  
  \caption{Generalized vortensity perturbation (relative to $t=0$) for
    cases of $\tilde{\beta}=0.1,1,10$ (left,middle,right) during
    the growth of non-axisymmetric modes. The planet potential has
    been switched off.  The number of vortices
    decrease as $\tilde{\beta}$ increases. Note that snapshots are
    taken later because it takes longer for the vortices to grow and
    become visible with increasing cooling time. 
    %    correspondingly increase to capture the longer growth phases of
    %    the instabilities. 
    \label{2Dlinear} {\bf I'm curious to know if plots clearer if
      perturbations are measured relative to t=30? remember t=30 is
      our reference state for instability. But we'll probably keep
      these for consistency (later non-linear plots are all measure
      with respect to t=0).}} 
\end{figure*}

@@

{\bf
If there's time, do low resolution planet off simulations but keep the planet on
until $t=40P_0$ or $50P_0$. Look at stability properties when the gaps
are generally deeper (including the nearly adiabatic case). Although
with the planet instability may already appear by $t=40P_0$, the
azimtuhal average performed prior to perturbations will erase it. 
}

@@

We now consider long term simulations of gap-edge vortices for
$\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
throughout.  
% up to a total of
%$2000P_0$. 
%Planets were left free to interact dynamiclly with the disk
%and vortices after $t=30P_0$ as apposed to previous section.  
We employ a grid with $(N_r,N_{\phi})=(512,1024)$ 
in order for these simulations to be computationally feasible. We also use a larger
disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimize boundary
effects on vortex evolution. We apply an open boundary at
$r_\mathrm{out}$. {\bf mention lower resolution preliminary runs -
  similar results to those presented here (?). want to demonstrate convergence}

@@

The linear growth of the RWI and vortex-formation is followed by 
vortex merging. For all $\tilde\beta$ values, only one vortex remains
by $150P_0$.
%i.e. only the $m=1$ mode remains. 
{\bf is vortex merging affected by $\tilde{\beta}$? e.g. timescale
  taken to reach one vortex as a function of cooling}
%by $150P_0$ for all , only $m=1$ vortex
%modes exist. 
The evolution of the amplitude of the $m=1$ mode in surface density is
shown in Fig~\ref{lifetimeplot} for   
different $\tilde\beta$. {\bf should state (also for planet-off runs)
  what the radial range is used for averaging the Fourier amplitudes}

@@

In all cases the system remains in a quasi-steady state for
$\gtrsim500P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex grows in intensity. As the vortex forms it has a
charcteristic value of $Ro\approx-0.1$ indicating anti-cyclonic
motion. As the $m=1$ amplitude grows the relative spin of the vortex
continuously grows along with it, reaching values of
$Ro\approx-0.4$, corresponding to surface density almost 10 times the
initial value. {\bf for what cooling are these values quoted for? do
  these values depend on cooling?} Vortex growth is attributed to
the continuous generation of vorticity by the planet-disc interactions. 
{\bf is the evidence for this the fact that growth is not observed in
  planet-off simulations?}

@@

Interestingly, in most cases we observe a sudden drop in 
the $m=1$ amplitude within the simulation time, which occurs over
timescales of $50P_0$, i.e. much shorter than the duration of the quasi-steady
state. {\bf is this dissipation timescale independent of cooling?}.  
%Eventually
%the $m=1$ amplitudes are also characterized by 
After these characteristic drops, the instability is never seen to
reform at such large overdensities and surface density bumps are never larger
than that due to the planetary wakes. We designate the time of these
amplitude drops as the lifetime 
of the quasi-steady vortex.  
{\bf question: does instability
  actually happen again? if it does, but just weaker, that may be
  because the first vortex has smoothed out the gap edge, so
  instabilities following that would be weaker.}

\begin{figure}
  \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
  \caption{Relative surface density perturbation for the
    $\tilde\beta=0.1$ case during quasi-steady state with a single
    vortex. The plot for other values of the cooling time $\tilde{\beta}$ are
    similar. {\bf is this plot truncated? in the text it says the
      domain goes to $r=45$?}
    %The large scale non-axisyemetric $m=1$ overdensity can be
    %seen. 
    \label{Vortex2D}} 
\end{figure}

@@

As the vortex dies off as described above its associated Rossby number
of quickly approaches $0$. Gradients of surface density are seen to
spike up in wake-like features around the vortex during dissipation. 
{\bf show
  this. e.g. vortex before, during and after death. maybe show
  snapshots of surface density perturbation (vortex) compared with
  $\nabla\Sigma/\Sigma$ (vortex shocks). demonstrate vortex shape
  changes and shocks develop prior ot death. state whether this is
  generic for all coolings}.  
In addition, we also observed large increases in the Mach number near
the vortex during dissipation, which indicate susceptibility to
shocks. {\bf show mach number plot, or quote values. relevant mach
  number is fluid flow speed with respect to 
  vortex bulk motion around the disc. does the typical mach number
  near the vortex just before its death depend on cooling?} 
Thus, the sudden drop in vortex amplitude may  
be a result of the development of associated spiral shocks dissipating
its energy \citep[cf.][who 
  suggested damping due to planet-induced shocks]{fu14}.

{\bf note that fu et al suggested vortex damping by the planet
  shocks. here, we are suggesting vortex dissipates energy by shocking
  the disk. our evidence is that we observe vortex growth before it
  dissipates, and spiral shocks associated with the vortex always
  develop before vortex death. once it starts shocking the disc it
  loses energy. if the planet-induced shocks were responsible, death
  should be more gradual. can we think of other evidences that it's
  not the planet shock that's responsible?                    
}

Vortex growth during the quasi-steady state also acts to smooth out
the gap edge. 
%Of note is also that the large overdensity the vortex
%creates during it's quasi-steady state distorts the background
%disk. 
{\bf show this (smoothening of gap edge due to vortex growth),
  e.g. general vortensity evolution or surface density 
  evolution}. 
This is because the vortex has an associated viscosity {\bf quote a
  typical value of alpha viscosity associated with vortex}. This acts
against gap-opening by the planet, so that the condition for the RWI
becomes less favourable as the vortex grows. {\bf self-limiting
  behaviour?}

\subsection{Vortex lifetimes as a function of cooling rate}

@@

Thus, unlike our `planet-off' simulations we find that there is
a cooling time $t_c$ which maximises the vortex lifetime. {\bf to make
this comparison we need to discuss vortex lifetimes for planet-off
simulations as well} 
This
non-monotonic behaviour was also reported by \cite{fu14}, who
performed isothermal disc-planet simluations with different fixed
aspectratios.   
%Cooling rates with
%$\tilde\beta>5.0$ showed lifetimes that were well below the max
%lifetime. 
By analyzing the region
around the vortex {\bf define this region} we find the
$\tilde\beta=5.0$ case, with the longest vortex lifetime, has
$h\approx0.06$. This value of $h$ was also found to maximize vortex
lifetimes in the isothermal simulations of \cite{fu14}.

>>>>

\section{Growth of non-axisymmetric modes without the influence of the
  planet}\label{linear1}
In this section, the planet is introduced at $t=20P_0$ and 
its potential switched on over $10P_0$. At $t=30P_0$ we switch off the
planet potential and azimuthally average the surface density, energy
and velocity fields. At this point the planet has carved a partial
gap and the RWI has not yet occured.
 We then perturb the surface density for $r>r_p$ and continue to 
evolve the disc. We impose  sinusoidal perturbations with 
azimuthal wavenumbers $m\in[1,10]$ and random amplitudes within $\pm 0.01$
 times the local surface density.% {\bf what are PERTMIN and PERTMAX values?}  
This procedure allows us to analyse the growth of 
non-axisymmetric modes associated with the gap, but without
complications from non-axisymmetry arising directly from disc-planet
interaction.

@@

Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
with an open boundry condition at the outer disk edge $r_\mathrm{out}=25r_\mathrm{in}$. We consider 
%for a time up to $200P_0$ allowing high resolution and detection of high
%$m$ structures.
cases of $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
and slowly cooled discs. % {\bf boundary condition for these sims?}

@@

The intial planet-disk interaction form bumps and grooves in the gap profiles
 which can potentially be unstable due to axisymmetric instabilities. The
 generalized local axisymmetric stability condition is the Solberg-Hoiland
 criterion,
\begin{equation}
 \kappa^2+N^2 \geq 0 $ where $ N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r} \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right)
\end{equation}
where $N$ is the Brunt-V\"ais\"al\"a frequency for the disk due to entropy
 variation which is non-zero for non-adiabatic models such as being considered.
 At  $t=30P_0$ the outer gap edge of $r=2.5r_h$,  where the RWI is excited the
 criterion value, is found to be approximately $0.002$ for all $\tilde\beta$
 which is the local minimum in the gap region which indicates local
 axisymmetric stability of gap. The Solberg-Hoiland criteria is similary
 satisfied for the entire disk and for all times within the simulation length
 with little change of values. Thus for all cases of $\tilde\beta$ the
 planet-induced gaps formed are stable to axisymmertic instabilities.

@@

In all cases we observe growth of non-axisymmetric structures after the
planet potential is switched off and the system subject to random
perturbations. An example is shown in Fig. \ref{linearmodes} for 
$\tilde{\beta}=10$. We charcterize these
modes with an azimuthal wavenumber $m$ and growth rate $\gamma(m)$ as defined by
Eq.~\ref{fouriertransform}---\ref{growth} with radial range of $[2r_h,5r_h]$. Table \ref{modetable}
lists the growth rates for the three cooling
times, measured during linear growth.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
    0cm 0cm 0cm]{figures/linear_stability}
  \caption{Evolution of azimuthal Fourier modes of disc surface
    density, non-dimenionlized by the axisymmetric background keplerian mode
 $\Sigma_0(t=0)$ for the
    `planet-off' simulations with $\tilde{\beta}=10$. Colours correspond
    to different $m$ values. The $m=3$ component is the fastest growing
    mode during linear growth and has a corresponding
    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
% {\bf is the $\Sigma_0$ at t=0?} }  
\end{figure}

@@

\begin{figure}
  \centering
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity50}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity67}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity75}
  }  
  \caption{Generalized vortensity perturbation (relative to $t=0$) for
    cases of $\tilde{\beta}=0.1,1,10$ (left,middle,right) during
    the growth of non-axisymmetric modes. The planet potential has
    been switched off.  The number of vortices
    decrease as $\tilde{\beta}$ increases. Note that snapshots are
    taken later because it takes longer for the vortices to grow and
    become visible with increasing cooling time. 
    %    correspondingly increase to capture the longer growth phases of
    %    the instabilities. 
    \label{2Dlinear} {\bf I'm curious to know if plots clearer if
      perturbations are measured relative to t=30? remember t=30 is
      our reference state for instability. But we'll probably keep
      these for consistency (later non-linear plots are all measure
      with respect to t=0).}} 
\end{figure}

@@

%{\bf
%If there's time, do low resolution planet off simulations but keep the planet on
%until $t=40P_0$ or $50P_0$. Look at stability properties when the gaps
%are generally deeper (including the nearly adiabatic case). Although
%with the planet instability may already appear by $t=40P_0$, the
%azimtuhal average performed prior to perturbations will erase it.}

@@

We now consider long term simulations of gap-edge vortices for
$\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
throughout.  
% up to a total of
%$2000P_0$. 
%Planets were left free to interact dynamiclly with the disk
%and vortices after $t=30P_0$ as apposed to previous section.  
We employ a grid with $(N_r,N_{\phi})=(512,1024)$ 
in order for these simulations to be computationally feasible. We also use a larger
disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimize boundary
effects on vortex evolution. We apply an open boundary at
$r_\mathrm{out}$.

Preliminary simulations with same numerical scheme but lower resolution of
 $(N_r,N_{\phi})=(256,512)$ where also done for the above $\tilde\beta$ values
with the inclusions of $\tilde{\beta}=0.25,0.75,2.5,7.5$ as well.
Results and vortex behaviour are convergent with
higher resolution simulations as discussed in this section.
 {\bf mention lower resolution preliminary runs -
  similar results to those presented here (?). want to demonstrate convergence}

@@

The linear growth of the RWI and vortex-formation is followed by 
vortex merging. Merging timescales are independent of $\tilde\beta$ value,
 and by $60P_o$ only one vortex remains as the growth of these modes are now
 accelerated by planetary influence.
%i.e. only the $m=1$ mode remains.

The evolution of the amplitude of the $m=1$ surface density mode
 over the radial range of $[2r_h,10r_h]$ is shown in Fig~\ref{lifetimeplot}
 for different $\tilde\beta$.
% {\bf should state (also for planet-off runs)
 % what the radial range is used for averaging the Fourier amplitudes}

@@

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex grows in intensity. As the vortex forms it has a
 value of $Ro\approx-0.1$ for all $\tilde\beta$
 indicating anti-cyclonic
motion. As the $m=1$ amplitude grows the relative spin of the vortex
continuously grows along with it, reaching values of
$Ro\approx-0.4$ characteristicly for all cooling rates. Relative surface
 density pertibations at the center of vortices can reach values of 7 times the
 intial density for all cases of $\tilde\beta$ in this quasi-steady state and
 can be found as high as 9 times for the $\tilde\beta=1.0$ case.
 The gap profiles apply a surface density boost in the vortex region as the
 vortices are located at the gap edge maxima
 but the values associated with this boost have relative density pertibations
 of under 1 and thus are negligibe compared to the max amplitudes of the vortices.
 Vortex growth is attributed to
the continuous generation of vorticity by the planet-disc interactions. 
This is supported by the evidence that in long term simulations where the
planet is removed such as in section \ref{nonlinearplanetoff} the amplitude of
 the vortices are not seen to contiunously grow without the planet influence.

{\bf is vortex merging affected by $\tilde{\beta}$? e.g. timescale
  taken to reach one vortex as a function of cooling}
{\bf for what cooling are these values quoted for? do
  these values depend on cooling?}
 {\bf is the evidence for this the fact that growth is not observed in
  planet-off simulations?}

@@

Interestingly, in most cases we observe a sudden drop in 
the $m=1$ amplitude within the simulation time, which occurs over
timescales of $t\in[20P_0,70P_0]$, i.e. much shorter than the duration of the
quasi-steady state. Dissipation timescales follow the basic trend of decreasing with increasing $\tilde\beta$ but more accurately correlated with Mach number values which will be discussed in following section
{\bf is this dissipation timescale independent of cooling? Trend is based on mach values which dont follow linear trend } 
%Eventually
%the $m=1$ amplitudes are also characterized by 
After these characteristic drops, the instability is never seen to
reform at such large overdensities and surface density bumps are never larger
than that due to the planetary wakes. We designate the time of these
amplitude drops as the lifetime 
of the quasi-steady vortex.
The instability is not seen to emerge again after the intial vortex dissipation
 due to the smoothening out of the gap edge during dissipation by the
 transfer of density from the once tightly held vortex region to the surrounding
 disk and gap.  
{\bf question: does instability
  actually happen again? if it does, but just weaker, that may be
  because the first vortex has smoothed out the gap edge, so
  instabilities following that would be weaker.}

\begin{figure}
  \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
  \caption{Relative surface density perturbation for the
    $\tilde\beta=0.1$ case during quasi-steady state with a single
    vortex. The plot for other values of the cooling time $\tilde{\beta}$ are
    similar.% {\bf is this plot truncated? in the text it says the
      %domain goes to $r=45$?}
    %The large scale non-axisyemetric $m=1$ overdensity can be
    %seen. 
    \label{Vortex2D}} 
\end{figure}

\subsection{Non-linear instability without planet influence} \label{nonlinearplanetoff}
Non-linear observations were also made for the `planet-off' simulations.
After the linear growth phase of the vortices, vortex merging takes hold but
 now on timescales of up to $150P_0$ as the vortices grow as linear
 instabilities discussed previously. The vortex merging time is depedendent on
 the growth rates of the modes and how long they take to saturate. Similar
 quasi-steady vortices form but the long-term behaviour is significantly
 different without the planet as seen in Fig.\ref{planetofflifetimeplot}.
 The quasi-steady vortices formed have significantly lower density pertibations
 and now slowly decrease in amplitute in time as apposed to previously
 where their growth was supported by planet interactions. The vortices simply
 dissipate viscously and no longer undergo the quick dissipation that
 characterized the dissipation as earlier. The dissipation rate is dependent
 on the cooling rate with viscous dissipation decreasing with larger
$\tilde\beta$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
    0cm 0cm 1.1cm]{figures/longterm_planetoff}
  \caption{Similar plot as Fig. \ref{lifetimeplot} but with `planet-off' simulations. Vortices dissipate viscously without planet driving influence} \label{planetofflifetimeplot}
\end{figure}

@@

Gradients of surface density are seen to
spike up in wake-like features around the vortex during dissipation. 
Snapshots of relative density pertibation and density gradients during the
quasi steady state, during disspation and just before death are displayed in
Fig. \ref{shockplot} for $\tilde\beta=1.0$. Quasi-steady vortices have large 
elliptical shapes, while dissipating vortices are denser and smaller in spacial
extent.
As the vortices grow in intensity wakes are seen to extend from them. These
wakes correlate with large gradients in surface density and are first seen
when vortices are in the later half of their quasi steady states. During this
phase the average value of the gradients along the shock region are
$|\nabla\Sigma| \approx 4 \Sigma$. Just before the vortices start dissipation
this quantity sharply increases to a value of $ \approx6 \Sigma$ in which it
 remains during dissipation and rest of vortex lifetime.  
As the vortex dies off as described above its associated Rossby number quickly
approaches $0$.

In addition, we also observed large increases in the Mach number near
the vortex during dissipation, which indicate susceptibility to
shocks. Mach numbers with repect to the vortex bulk motion around the vortex
 region of $|r-r_{vortex}|<2H$, where $H$ denotes the gas scale height, is
 $\approx 0.8$ after the intial formation of the vortices for all
 $\tilde\beta$. During the quasi-steady state the Mach number increases at a
 steady rate with larger increases for lower $\tilde\beta$. Time values in
 which Mach numbers reach values of 1 approximately correlate with the start 
of dissipation and during dissipation values quickly increase to larger max values.
 Dissipation timescales correlate with 
Mach numbers reached during dissipation as the quickest dissipation occurs for 
$\tilde\beta=0.5$ which is on order of $70P_0$ and reaches Mach number
 $\approx1.2$ while the $\tilde\beta=0.1$ case which dissipates on scales o
f $20P_0$ reaches Mach numbers $\approx1.4$
Thus, the sudden drop in vortex amplitude may  
be a result of the development of associated spiral shocks dissipating
its energy \citep[cf.][who 
  suggested damping due to planet-induced shocks]{fu14}.

{\bf note that fu et al suggested vortex damping by the planet
  shocks. here, we are suggesting vortex dissipates energy by shocking
  the disk. our evidence is that we observe vortex growth before it
  dissipates, and spiral shocks associated with the vortex always
  develop before vortex death. once it starts shocking the disc it
  loses energy. if the planet-induced shocks were responsible, death
  should be more gradual. can we think of other evidences that it's
  not the planet shock that's responsible?}
{\bf show
  this. e.g. vortex before, during and after death. maybe show
  snapshots of surface density perturbation (vortex) compared with
  $\nabla\Sigma/\Sigma$ (vortex shocks). demonstrate vortex shape
  changes and shocks develop prior ot death. state whether this is
  generic for all coolings}. 
{\bf show mach number plot, or quote values. relevant mach
  number is fluid flow speed with respect to 
  vortex bulk motion around the disc. does the typical mach number
  near the vortex just before its death depend on cooling?}

Vortex growth during the quasi-steady state and dissipation also acts to smooth
out the gap edge. To characterize the gap, we azimuthally and radially average
the relative density pertibation within $r\in[r_{planet},r_{out}]$ where
 $r_{out}=$ is defined as the outer gap radius defined as point where
$\langle\Delta \Sigma(r_{out})/\Sigma(t=0)\rangle_{\phi}=0$. Thus a more 
negative gap depth parameter correlates with a steeper gap edge. In
 Fig. \ref{gapdepth} the gap depth parameter with respect to simulation time
for $\tilde\beta=1.0$ is shown. The gap depth is seen to slowly increase as
vortex reaches its max amplutude at $t=1000P_0$ which is followed by a sharp
increase which correlates with the time of vortex dissipation.
%Of note is also that the large overdensity the vortex
%creates during it's quasi-steady state distorts the background
%disk.  
The smoothening out of the gap edge during later half of quasi steady state
 is because the vortex has an associated viscosity $O(10^{-2})$.
%{\bf quote a typical value of alpha viscosity associated with vortex}.
 This acts
against gap-opening by the planet, so that the condition for the RWI
becomes less favourable as the vortex grows.
The sudden smoothening of gap egde during dissipation is due to the large
influx of density from the vortex to the gap.
{\bf self-limiting
  behaviour?} 
{\bf show this (smoothening of gap edge due to vortex growth),
  e.g. general vortensity evolution or surface density 
  evolution}

\begin{figure}
\subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock1}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock2}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock3}
  } \\[-0.98cm]
\subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock4}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock5}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock6}
  } 
  \caption{Relative density pertibation of vortices (top) and density gradient
 (bottom) for before (left), during (middle) and after dissipation (right).
 Large density gradients are characteristics of shocks and become prominant
 around vortex during dissipation} \label{shockplot}
\end{figure}

\begin{figure}
    \includegraphics[width=\linewidth]{figures/gapdepth}
 \caption{Running average of relative density pertibation within the outer half
of gap} \label{gapdepth}
\end{figure}
\subsection{Vortex lifetimes as a function of cooling rate}

@@

Thus, unlike our `planet-off' simulations we find that there is
a cooling time $t_c$ which maximises the vortex lifetime. In the long term
 'planet-off' simulations we find that the vortices dissipate very gradually and visciously. {\bf to make
this comparison we need to discuss vortex lifetimes for planet-off
simulations as well} 
This
non-monotonic behaviour was also reported by \cite{fu14}, who
performed isothermal disc-planet simluations with different fixed
aspectratios.   
%Cooling rates with
%$\tilde\beta>5.0$ showed lifetimes that were well below the max
%lifetime. 
By analyzing the region
around the vortex {\bf define this region} we find the
$\tilde\beta=5.0$ case, with the longest vortex lifetime, has
$h\approx0.06$. This value of $h$ was also found to maximize vortex
lifetimes in the isothermal simulations of \cite{fu14}.
